\bf Model Visualization Techniques for a Social Network Model

### 2017-10-03 paper.Rnw
make the paper a child for the whole shebang

<<setup, echo = FALSE, message = FALSE, warning = FALSE>>=
options(replace.assign=TRUE,width=70, digits=3)
require(knitr)
opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.pos='h', out.width='.99\\textwidth', par=TRUE, cache=FALSE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE, dev="cairo_pdf", fig.width = 6, fig.height = 6)#, root.dir = "~/Desktop/Dissertation/SAOM-removing-blindfold/")
@

<<pkgs>>=
library(tidyverse)
library(RSiena)
library(network)
library(sna)
library(geomnet)
library(GGally)
library(netvizinf)
library(RColorBrewer)
library(gridExtra)
library(cowplot)
library(extrafont)
loadfonts(quiet = T)

ThemeNoNet <- theme_bw() %+replace% 
            theme(plot.title = element_text(size = 12,
                                            face = 'plain', 
                                            angle = 0, 
                                            family="Times New Roman"), 
                  axis.title.x = element_text(size = 12,
                                            face = 'plain', 
                                            angle = 0,
                                            family="Times New Roman"),
                  axis.title.y = element_text(size = 12,
                                            face = 'plain', 
                                            angle = 90,
                                            family="Times New Roman"),
                  axis.text.x.top = element_text(size = 12,
                                            face = 'plain', 
                                            angle = 0,
                                            family="Times New Roman"), 
                  axis.text.x.bottom = element_text(size = 12,
                                            face = 'plain', 
                                            angle = 0,
                                            family="Times New Roman"), 
                  axis.text.y.left = element_text(size = 12,
                                            face = 'plain', 
                                            angle = 0,
                                            family="Times New Roman"),
                  axis.text.y.right = element_text(size = 12,
                                            face = 'plain', 
                                            angle = 0,
                                            family="Times New Roman"),
                  strip.text.x = element_text(size = 12,
                                            face = 'plain', 
                                            angle = 0, 
                                            family="Times New Roman",
                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.text.y =element_text(size = 12,
                                            face = 'plain', 
                                            angle = 90,
                                            family="Times New Roman",
                                            margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt")),
                  strip.background = element_rect(colour = "black", fill = "white")
                    )
ThemeNet <- theme_net() %+replace% 
            theme(plot.title = element_text(size = 12,
                                            face = 'plain', angle = 0, family = "Times New Roman"), 
                  strip.text.x = element_text(size = 12,
                                            face = 'plain', angle = 0, family = "Times New Roman", 
                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.text.y =element_text(size = 12,
                                            face = 'plain', angle = 0, family = "Times New Roman",margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.background = element_rect(colour = "black", fill = "white"))
@

\section{Introduction} 
% \st{
% \begin{itemize}
% %\item What? - model visualization and SAOMs. briefly dsecribe what these things are
% %\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks. 
% \item How?
% \end{itemize}
% %also talk about other dynamic network models briefly to put SAOMs in context.
% }
%\hh{XXX - talk a bit about social networks first, why we want to model them, and that usually the models are not dynamic.}
\st{%XXX - okay, here is what i have done for that:
Social networks have been studied for decades, beginning with a few foundational works, including the 1967 study, "The Small World Problem" by Stanley Milgram \citep{goldenberg09}. Examples of social networks include collaboration networks between academic researchers, friendship networks in a school or university, and trade networks between nations. In recent years, the study of social networks has grown in popularity due to an increase in the availability and access to social network data. There are many kinds of social networks, but there are not as many statistical models for social network data.} Some network models that have been applied to social networks include the exponential random graph model and latent space models. These models, however, are only for single instance networks. If we only have one network observation, or only care about one state of a complex network in time, like a snapshot of the World Wide Web, using the well-established models for single network observating is not a problem. If, on the other hand, we have many observations of social network over time, these models may not be appropriate because they do not explicitly allow for the network to change as time passes. When studying a network over time, referred to as a \textit{dynamic network}, we need a model that can take the time aspect of the network into account. 
Models for dynamic social networks have a great deal of modelling potential because of how realistic their structure can be. A social network does not form spontaneously: it evolves over time. Ties are formed and dissolved, and new actors join the social structure. Modeling the underlying mechanisms that create network changes over time is very complex but also provides potential to uncover hidden truths.

\par In this paper, we concentrate on one type of model for dynamic social networks: the stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}. These models are fundamentally different from other social network models because they allow us to incorporate network \textit{and} actor statistics, where other models only rely on the network statistics, to model the changes in the network. Allowing the actor-level statistics to directly effect the structure of the network leads to a more practical and relevant approach to model change in a social network. In the ``real" world we expect people with common interests to be more likely to form relationships, and SAOMs allow us to incorporate this intuition in the modeling process.

\st{
%\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}. This software marks a huge contribution to the field of social network analysis., and when we began working with it we were eager to understand how it works. HH: we are still interested in this.
%\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood? I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g. likelihood functions turn into very complex objects because of the sheer number of parameters involved. Therefore, computationally more tractable solutions are used to fit estimators, in particular, SAOMs are fit using ... }
Unlike other network models, SAOMs are not very well understood. They are relatively new, especially compared to the classic exponential random graph models, and they are not very tractable analytically. Likelihood functions quickly very complex objects to analyze due to the dependency structure inherent in the data. Therefore, computationally more tractable solutions are used to fit estimators, and in particular, % Traditional modeling steps, like determining the likelihood function, are very complex. Because of this, 
SAOMs are often fit to data using a series of Markov Chain Monte Carlo (MCMC) phases for finding method of moments estimators. In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, which was developed by \citet{RSiena}. This software marks a huge contribution to the field of social network analysis, but the many moving pieces involved in parameter estimation are largely ``behind the scenes" and hidden from the software user.} In this paper, in order to better understand the model-fitting process, we attempt to bring SAOM fits and the fitting process out of their black boxes, by combining the principals of network visualization with those of model visualization as discussed in \citet{modelvis}. %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
By bringing some light to the underlying methods and structures that are behind the scenes when fitting SAOMs to network data, we aim to help researchers working with the models better understand the implications and analyses of these models.

\par We get into the SAOM black box by using model visualization (see \citet{modelvis}) to display the model in the data space, view collections of models instead of single models, and visually explore the process of fitting the SAOMs as opposed to looking at only the final output. Stochastic actor-oriented models are a prime example of a set of models that can benefit greatly from the application of model visualization. For instance, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden from the analyst in the model fitting process. Bringing the CTMC out of the black box and into the light through model visualization can provide researchers with insights into the underlying features of the model. Furthermore, SAOMs can include a great deal of parameters to be added to the model structure, each of which is attached to a network statistic. These statistics are often somewhat, if not highly, correlated, which causes high correlation between the associated parameters in a SAOM. By visualizing collections of SAOMs, we gain a better understanding of these correlations and find ways to deal with them and rectify their effects in the model. In addition, the estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and the convergence checks for these estimates rely on simulation from the fitted model. Again, each of these steps are largely kept in the background of the estimation process. With the help of static and dynamic visualizations we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.

In Section~\ref{sec:netintro}, we introduce basic concepts of networks and network visualizations. In Section~\ref{sec:saoms}, we present the family of stochastic actor-oriented models for social network analysis. In Section~\ref{sec:ModelVis}, we combine concepts from Sections~\ref{sec:netintro} and \ref{sec:saoms} in an application of the model-vis paradigm, and conclude with a discussion in Section~\ref{sec:discussion}.

\section{Networks and their Visualizations} \label{sec:netintro}

\subsection{Introduction to Network Structures}

Network data is of frequent interest to researchers in a wide array of fields. There are technological networks, like power grids or the internet, information networks, such as citation networks or the World Wide Web, biological networks, like neural networks, and social networks, just to name a few \citep{newman}. Each of these examples have one thing in common: their data \textit{structure}. There are always units of observation: the power stations, websites, neurons, and people, which we refer to throughout this paper as \textit{nodes} or \textit{actors}. There are also always connections of some kind between those units: the power lines, hyperlinks, electrical signals, and relationships, which we will call \textit{edges} or \textit{ties}. Networks might change over time, like when new websites and hyperlinks are added on the World Wide Web, or when there are new people and relationships in a friendship network. The nodes and edges themselves can also have inherent variables of interest, e.g.\ the institution of authors in a co-authorship network, or the number of times two authors have collaborated.

The multiple layers of network data structures pose unique problems to network analysts. Some questions that network researchers may aim to answer are: How does the strength of a tie between two nodes affect the overall structure of the network? Do node-level differences affect the formation or dissolution of edges? Which views of the data are most informative for communicating significant effects and other results of statistical analyses of the network of interest? These are, of course, just a few broad questions, and we focus here on the latter, which we aim to answer through visual exploration of network data and models.

\subsection{Visualizing Network Data} \label{sec:netvizintro}

Network visualization, also called network mapping, is a prominent subfield of network analysis. Visualizing network data is uniquely difficult because of the structure of the data itself.  Most, if not all, data visualizations rely on well-defined axes inherited from the data. If variables are numerical, histograms, scatterplots, or time series plots are straightforward to construct. If the variables are categorical, bar charts and mosaic plots are available to to the researcher. If the data are spatial, there is a well-defined region in which to view information. Network data, however, are much less cut-and-dried.

There are two primary methods used to visualize networks: node-link diagrams and adjacency matrix visualization \citep{knuth2000, adjmatpro}. As a toy example, let us assume that we have five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$ We use this toy data set to demonstrate the two visualization methods in Figure~\ref{fig:toyplots}.

<<toyex, echo = FALSE>>=
set.seed(81393)
nodes <- data.frame(id = 1:5, group = sample(LETTERS[1:2], size = 5, replace = T), stringsAsFactors = F)
g <- igraph::as_data_frame(igraph::random.graph.game(n = 5, p = .4))
toynet <- list(nodes = nodes, edges = rbind(g, c(5,4)))
@

The first method, the node-link diagram, represents nodes with points in 2D Euclidean space and then represents edges by connecting the points with lines when there is an edge between the two nodes. These lines can also have arrows on them indicating the direction of the edge for directed networks. But because there is typically no natural placement of the points unless they have important spatial locations, a random placement of the points is used, then adjusted via a layout algorithm, of which there are many \citep{drawingalgs}.

<<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our directed toy network, with nodes placed using the Kamada-Kawai algorithm. On the right, the adjacency matrix visualization for that same network.", out.width='.49\\textwidth'>>=
toynetdat <- fortify(as.edgedf(toynet$edges), toynet$nodes)
ggplot(data = toynetdat) + 
  geom_net(aes(from_id=from_id, to_id=to_id), labelon = T, size = 11, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, fontsize = 9, arrowgap = .05, directed = T) +
  ThemeNet +
  theme(plot.title = element_text(size = 20)) + 
  labs(title = "Node-Link Diagram")
#ggsave("figure/toynetnld.pdf",plot = nld, width = 4, height = 4)

ggplot(data = toynet$edges) +
  #geom_tile(aes(x = from, y = -to), color = 'grey20', fill = 'black') +
  geom_tile(aes(x = to, y = -from), color = 'grey20', fill = 'black') +
  geom_segment(aes(x = x, xend = xend), y = -0.5, yend=-5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) + 
  geom_segment(aes(y=-y, yend=-yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = -0:5+.5, yend= -0:5+.5)) + 
  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5), position = "top") +
  scale_y_continuous(breaks = (-5):(-1), labels = 5:1, limits = c(-5.5,-.5)) +
  ThemeNoNet + 
  theme(panel.grid = element_blank(), aspect.ratio = 1, 
        plot.title = element_text(size = 20), 
        axis.title.x.top = element_text(size = 18),
        axis.title.y.left = element_text(size = 18), 
        axis.text.x.top  = element_text(size = 18),
        axis.text.y.left  = element_text(size = 18)) +
  labs(title = "Adjacency Matrix Visualization", x = "\nto (alter)", y = "from (ego)\n")
@
%\hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }

Some commonly used layout algorithms, such as  the Kamada-Kawai layout \citep{kamadakawai} and the Fruchterman-Reingold layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them. In these algorithms, the edges of the network act as springs pushing and pulling the nodes in a low dimensional (usually two-dimensional) space. Another algorithm uses multi-dimensional scaling, relying on distance metric and computing a matrix whose entries represent the ``distance" between every pair of nodes. There are also layout algorithms that use properties of the adjacency matrix, like its eigenstructure, to place the nodes in 2D space \citep{drawingalgs}. The node-link diagram using the Kamada-Kawai layout algorithm for our toy network is shown in Figure~\ref{fig:toyplots}. Unless otherwise stated, all other node-link diagrams in this paper will use the Kamada-Kawai layout. %Other algorithms use properties of the network's adjaceny matrix in order to place the nodes.
%hh{XXX how does the adjacency matrix feature here? The adjacency matrix is a different mathematical representation of a network, but describes the same network as a node-link diagram.} \st{XXX I'm taking this to mean I should describe adjacency matrices more, so I've added a discussion below. XXX }

\st{The second primary method for network visualization uses the adjacency matrix of the network. The adjaceny matrix of a network, $\mathbf{A}$, describes the edges of a network in matrix form. An entry $A_{ij}$ of $\mathbf{A}$, for two nodes $i\neq j$ in the network is defined as 
\[
  A_{ij} =
  \begin{cases}
                                   1 &  \text{if an edge exists } i\to j \\
                                   0 & otherwise
  \end{cases}
\]
Note here that our edge variables are binary: we only consider the presence or absence of an edge. If the network has weighted edges, for example an email network where edge weights represent the number of emails sent from one person to another, the entries in the adjacency matrix are the edge weights instead of zeroes and ones. In an undirected network, $A_{ij} = A_{ji}$, but in a directed graph this is only true if there is an edge from $i$ to $j$ and from $j$ to $i$. Thus, $\mathbf{A}$ is always symmetric for undirected networks, and is symmetric for directed networks only if every edge between two nodes is reciprocated.}
%\hh{XXX what is an adjacency matrix? - explain at this point in mathematical terms} \st{XXX see above XXX}
An adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}.

Each type of visualization comes with its own advantages and disadvantages. For example, paths between two nodes in a network are easier to determine with node-link diagrams than with adjacency matrix visualizations \citep{adjmatviz}. In node-link diagrams, node-level information can be incorporated into the visualization by coloring or changing the shape of the points representing the nodes, and edge-level information can be incorporated by coloring the lines, or changing their thickness, linetype, or color. Incorporating a node-level variable into an adjacency matrix visualization is not as straightforward or simple, which is more focused on edges. Adjacency matrix visualization has been found to be particularly useful when the network is very complex, dense, or large, and experimental studies have shown adjacency matrix visualization to be superior to node-link diagrames for large networks. For example, for basic perceptual tasks on networks, including node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}. One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$: the edge $x_{ij}$ for $i \neq j$ appears in $\mathbf{A}$ twice: in $A_{ij}$ and $A_{ji}$, and so it also appears twice in the adjacency matrix visualization.  %\hh{XXX What is the exact logic here that Ghoniem used to come to the conclusion that edges in undirected graphs are over-represented? Why would directed graphs make that better? wouldn't that just add to the number of edges? } \st{XXX in undirected nets, the adj mat is symmetric, where one edge ($x_{ij}$) is represented in entry $i,j$ in the matrix and in entry $j,i$. So, there are many repetitions in the matrix, which makes the network appear more dense than it really is.} 
\st{This, however, may actually be an advantage for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, due to the fact that the edges $x_{ij}$ and $x_{ji}$ are not interchangeable. A node-link diagram, however, may underrepresent the edge count if the edges $x_{ij}$ and $x_{ji}$ both exist and are drawn on top of one another. Ultimately, however t}here is not one "correct" way to visualize network information, and we will be using both \st{the node-link and adjacency matrix visualization} methods throughout this paper to explore social networks and stochastic actor-oriented models.

\section{Stochastic Actor-Oriented Models for Longitudinal Social Networks} \label{sec:saoms}

% \st{What do I need to communicate in this section? 
% \begin{itemize}
% \item What the models are for: dynamic social networks with actor-level covariates
% \item What are the model components? rate function, objective function (network statistics, actor covariates, interactions between covariates and network statistics), time points ($t_m = 1, \dots, M$)
% \item What is RSiena? What methods of estimation, convergence checks does it use? What are the algorithms? What kinds of models can it fit? 
% \end{itemize}}

%The stochastic actor-oriented models are very aptly named: they model dynamic networks that can account for the individual node characteristics when making the edge changes. The object of analysis is a social network that has been observed at several discrete time points. The nodes in the network can have associated covariate variables that may be incorporated into modeling the tie changes in the network. The model itself has two primary pieces: the rate of change of ties, and the objective function of the nodes, which determines the exact change that gets made.

%We start with a set of network observations at $M$ discrete time points: $x(t_1), x(t_2) \dots, x(t_M)$. The first observation, $x(t_1)$ is always conditioned on. The number of nodes / actors / vertices is constant in all time points and is denoted $n$.  The tie variables are directed and  binary, and are denoted at $x_{ij}(t_m)$ for time point $m$, and this is sometimes abbreviated to $x_{ij}$ in notation. If there is a tie from node $i$ to node $j$, then $x_{ij}=1$, and $x_{ij}=0$ otherwise.

%\hh{mostly lit review plus a detailed description of the model and fitting process.}

A Stochastic Actor-Oriented Model (SAOM) is a model that incorporates all three components of dynamic networks: edge, node, and time information. It models the change of a network over time, allowing for changes in network structure due to actor-level covariates. This model was first introduced by Snijders in 1996 \citep{saompaper}. The two titular properties of SAOMs, stochasticity and actor-orientation, are crucial to understanding networks as they exist naturally. Most social networks, even holding constant the set of actors over time, are ever-changing as relationships decay or grow in seemingly random ways, and most actors (or nodes) in social networks have inherent properties that could affect how they change their role within the network, and vice versa.

\subsection{Definitions, Terminology, and Notation}

In this paper, the term \textit{dynamic network} refers to a network, consisting of a fixed set of $n$ nodes, that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 \le t_2 \le \dots \le t_M$. We denote the network observation at timepoint $t_k$ by $x(t_k)$. In the modelling process, we condition on the first observation, $x(t_1)$. The SAOM assumes that this longitudinal network of discrete observations is embedded within a continuous time Markov chain (CTMC), which we will denote $X(T)$. This process is almost entirely unobserved: we assume that the beginning of the process, $X(0)$, is equivalent to the first network observation $x(t_1)$, while the end of the process,$X(\infty)$, is equivalent to the last observation $x(t_M)$. Nearly all other parts of the process are unseen, with the exception of $x(t_2), \dots, x(t_{M-1})$. Unlike the first and last observations of the network, these ``in-between" observations do not have direct correspondence with steps in the continuous time Markov chain. Thus, the ``in-between" observations are considered to be ``snapshots" of the network at some point between two steps in the CTMC. The whole process $X(T)$ is a series of single tie changes that happen according to some pre-defined rate function, where one actor at a time is given the opportunity to add or remove one outgoing tie, or to not make any changes. Once an actor is chosen at random according to the rate function, it is ``given" the chance to change a tie, and it tries to maximize its utility function based on the current and near future states of the network. We expand on the model description further in the subsequent sections.

\subsubsection{The Rate Function}

For the network $x$ and each actor $i$ in the network, the number of times that an actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network is dictated by a \emph{rate function} $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$. For this paper, we assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant across all actors between observations at time $t_m$ to $t_{m+1}$, \st{thus our rate function is just a rate parameter in the overall model}. In general, SAOMs can incorporate covariate values and network statistics into the model, so that each node will have a different rate of change.  %\hh{XXX that means that the function is essentially just a rate parameter, right?} \st{XXX yes XXX}
Other modelling scenarios allow this rate to be more flexible,  e.g.\ a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics. 
%\hh{How is the rate parameter different from a rate parameter?} \st{I think you meant how is a rate fn different from a rate parameter? I updated the language XXX->}%In our model with the simple rate function, 
\st{In our simple model with a simple rate parameter instead of a rate function,} the rate parameter dictates how quickly an actor $i$ gets an opportunity to change one of its ties to the other nodes in the network, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$. If $j = i$, no change in the network is made. The model also assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state. \st{Let $\tau(i|x,m) $ be the wait time until actor $i$ makes its next change from its current state in the network $x$. Note that $m$ indicates the number of the wave that is conditioned on in the SAOM. For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$. The conditional independence assumption is expressed in Equation~\ref{eq:c1}.

\begin{equation} \label{eq:c1}
 \tau(i|x,m) | x_{i1}(m), \dots, x_{in}(m) \stackrel{\text{iid}}{\sim} Exp(\alpha_m)
\end{equation}

 The waiting time to the next change opportunity by \textit{any} actor in the network is also exponentially distributed with expected value $(n\alpha_m)^{-1}$. The distribution of waiting time for the whole network to change, $\tau(x|m) = \sum_i \tau_i(m) |x_{i1}(m), \dots, x_{in}(m) $ can then be written as

\begin{equation} \label{eq:c2}
 \tau(x|m) \sim Exp(n\alpha_m)
\end{equation}

} The parameter for the wait time for the whole network $n\alpha_m$ is the rate at which any tie change occurs. The estimation of this parameter is straightforward: a the method of moments is used to estimate the rate with the statistic $$C = \sum_i\sum_j |x_{ij}(t_{m+1}) = x_{ij}(t_m)|$$ which is the total number of changes from observation at time $t_m$ to the observation at time $t_{m+1}$. %In order to achieve the memorylessness property of a Markov process,

% \hh{XXX we tried to estimate the rate based on the simulations - you could include a histogram of the number of changes for the 1000 simulations in the viz part XXX}
% \st{XXX see line 1257 for plot. where should I put it? }
% \hh{XXX Put a reference to the plot here.}

\subsubsection{The Objective Function}

%\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
Because of the conditional \st{in}dependence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is \st{allowed to change} at a time. \st{The node $i$, which is the node that is chosen to change at the current time point, is called the \textit{ego} node. It has the potential to interact with all other nodes in the network, $j \neq i$. These nodes $j$, are referred to as \textit{alter} nodes, or simply \textit{alters}. These nodes are acted upon by the ego node, and they only act when they become the ego node at a subsequent time point in the CTMC.} For the ego node, $i$, in the current network state $x$, its objective function, which it tries to maximize, is written as 
\begin{equation} \label{eq:of1}
f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),
\end{equation}

for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$, the matrix of covariates. The vector $\boldsymbol{\beta}$ contains the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$. %\hh{XXX ego node isn't defined yet - the description of the charts would be a good place for it.} \st{XX I added it above the obj function equation. Is it ok there?} 
Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created thatis currently $x_{ij}=0$, or no change will occur.

The parameters, $\boldsymbol{\beta}$, correspond to various actor-level network statistics, $s_{ik}(x)$. %\hh{'There are always' ... that seems to come from a Snijder paper? Re-phrase this to be less vague: In the current implementation of RSiena (snijder) a model has minimally two parameters...} 
According to \citet[p. 371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node. \st{These effects should seem familiar to readers used to working with the classical exponential random graph model (ERGM) for networks. The outdegree represents the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the "rich get richer" effect), and the reciprocity parameter measures the tendency of outgoing ties to be returned within a network. The statistics corresponding to these effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$.} In the \texttt{RSiena} software that we use to fit the SAOMs, there are over 80 possible parameters to add to the model. The formulas for the effects are provided in \citet{RSienamanual}. The parameters, $\beta_k$, in the model can be split up into two groups: first, the structural effects, whose estimation depends only on the structure of the network, like the outdegree and reciprocity parameters mentioned above. The parameters are included when the researcher hypothesizes that they will model underlying mechanisms of network change. They hope to answer questions such as, ``How does the existing network structure influence change in the network?" %\hh{XXX move the last colored section to the front of this paragraph. That sets up the structure nicely, e.g. `In determining the covariates of a SAOM, two sources of parameters are considered: ...} \st{XXX as I'm going through and uncommenting your comments (sorry....) this one has lost meaning a little bit. Could you give me more detail? I'm not sure which section is which at this point. (sorry again...)} \hh{I guess you addressed this comment very thoroughly :)}
% and ``How do the behavior and characteristics of the nodes influence change in the network?" put this sentence below somewhere
The second set of effects are referred to as the actor-level or covariate effects. These covariate effects also depend on the structure of the network, with the additional inclusion of node-level covariates of interest. The covariate effects are written in terms of the tie variables $x_{ij}$, but also in terms of the covariates, $\mathbf{Z}$. A table of some possible structural and covariate effects is given in Table~\ref{tab:effects}. For a complete list of the network and covariate statistics that can currently be included in the objective function, see \citet{RSienamanual}.

\begin{table}
\centering
\begin{tabular}{m{4.5cm}m{4.5cm}m{3cm}}
\multicolumn{3}{l}{\textbf{Structural Effects}} \\ \hline\hline
outdegree  & $s_{i1}(x) = \sum_j x_{ij}$ & \includegraphics[width=2cm]{img/outdegree}\\ \hline
reciprocity  & $s_{i2}(x) = \sum_j x_{ij}x_{ji}$ & \includegraphics[width=2cm]{img/reciprocity}\\ \hline
transitive triplets  & $s_{i3}(x) = \sum_{j,h} x_{ij}x_{jh}x_{ih}$ & \includegraphics[width=1.5cm]{img/transtrip}  \\
\\
\multicolumn{3}{l}{\textbf{Covariate Effects}}\\  \hline \hline
covariate-alter  & $s_{i4}(x) = \sum_j x_{ij}z_j$ & \includegraphics[width=3cm]{img/covaralter} \\  \hline
covariate-ego  & $s_{i5}(x) = z_i\sum_j x_{ij}$ & \includegraphics[width=3cm]{img/covarego} \\  \hline
same covariate & $s_{i6}(x) = \sum_j x_{ij} \mathbb{I}(z_i = z_j)$ & \includegraphics[width=3cm]{img/covarsim}%\\
%jumping transitive triplets  & $s_{i7}(x) = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$
\end{tabular}
\caption{\label{tab:effects} Some of the possible effects to be included in the stochastic actor-oriented models in RSiena. There are many more possible effects, but we only consider a select few here. For a complete list, see the RSiena manual \citep{RSiena}.}
\end{table}

When node $i$ is given the chance to change a tie, %\hh{`we assume that they wish' sounds strange - don't assume wishes, state assumptions}
it attempts to maximize the value of its objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$, to account for unknown attraction between nodes. In \citet{modelsSnijders}, it is recommended that the $U_i(x)$ be random draws from a type 1 extreme value distribution. The additional random element is included to account for any random, unexplainable change in the network ties. % XX What is the thought behind that? } ``the type 1 extreme value distribution (or Gumbel distribution) with mean 0 and scale parameter 1" \citep[p. 368]{snijders01}. 
This distribution, which is also known as the log-Weibull distribution, has probability distribution function, using $\mu$ for the mean parameter and $\sigma$ for the scale parameter, of

\begin{equation} \label{eq:dist1}
g(u|\mu, \sigma) = \frac{1}{\sigma}\exp\left\{-\left(\frac{u-\mu}{\sigma} + \exp^{-\frac{u-\mu}{\sigma}}\right)\right\}.
\end{equation}

<<logweibull, fig.align='center', out.width = '.5\\textwidth', fig.height = 4, fig.cap='The probability distribution function for the type 1 extreme value distribution, also known as the log-weibull or Gumbel distribution with location parameter $ \\mu = 0$ and scale parameter $ \\sigma = 0$.'>>=
library(reliaR)
x <- seq(-3,10,.01)
y <- dgumbel(x = seq(-3,10,.01), sigma = 1, mu=0)
qplot(x = x, y= y, geom = "line") + 
  annotate('text', x = 6, y = .25, 
        label = "g(u) = exp(-(u + exp(-u)))",size=5, family = "Times New Roman")  + 
  theme(text = element_text(size = 20, family = "Times New Roman")) + 
  labs(x = "u", y = "density of g(u)")
@

The probability density function for the type 1 extreme value distribution is shown in Figure~\ref{fig:logweibull}. Using the distribution given in Equation~\ref{eq:dist1} with mean $\mu = 0$ and scale $\sigma = 1$ as \citet{modelsSnijders} suggests %\hh{XXX I'm not following - which distribution do you mean by `this' distribution? The density above is the type 1 extreme value density. I got lost at the  ``objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$"} \st{XXX I hope it is clearer now above}
is convenient because it \st{leads to a simple probability formula for} the probability that actor $i$ chooses to change its tie to actor $j$ \st{that can be written \textit{only}} in terms of the objective function. Let $p_{ij}(\boldsymbol{\beta}, x)$ be the probability that actor $i$ chooses to change its tie to actor $j$. Next, we write the network $x$ in its potential future state, $x(i \leadsto j)$, where the tie $x_{ij}$ has changed to $1-x_{ij}$. Then, the probility that the tie $x_{ij}$ changes is

\begin{equation}\label{eq:pij}
p_{ij}(\boldsymbol{\beta}, x) = \dfrac{\exp\left\{f_i(\boldsymbol{\beta}, x(i\leadsto j))\right\}}{\sum_{h \neq i} \exp\left\{f_i(\boldsymbol{\beta}, x(i \leadsto h))\right\}}
\end{equation}

When $i = j$ in $p_{ij}$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state. When the value of the objective function is high at the current state, the probability of not making a change in a microstep is also high. In the CTMC, when actor $i$ may make a change, it chooses which tie $x_{i1, \dots, x_{in}}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x)$. \st{The objective function and the resulting values of $p_{ij}$ are combined with the rate function to fully describe the CTMC that is used to model network change in a SAOM.}

\hh{XXX This paragraph ends a bit abruptly - what is the future state probability used for?} \st{XX The future state probability is used within the continuous time markov chain to move through the different network states. This needs more fleshing out. Adding to the to-do list.}

\subsubsection{Continuous Time Markov Chain (CTMC)}
\st{
In the continuous-time Markov chain literature, see for instance \citet{ctmc}, chains are characterized by their \textit{generator} or \textit{intensity} matrix $\mathbf{Q}$. This matrix describes the rate of change between two states of the CTMC process, and the rows of this matrix always add to zero. For directed networks with binary edge variables like the ones we will be working with, there are a very large number of possible states, $2^{n(n-1)}$: there are two possible states for an edge, and there are $n(n-1)$ edge relationships (excluding self-ties).  The intensity matrix for a CTMC in a SAOM is then a square matrix of dimension $2^{n(n-1)} \times 2^{n(n-1)}$. Only one tie changes at a time in the CTMC, resulting in $n(n-1)$ reachable states from the current network state. Thus, the intensity matrix $\mathbf{Q}$ is very sparse, with only $n(n-1) + 1$ non-zero entries in each row. Note that $n(n-1)$ of these entries represent the possible states that are one edge different from a given state, while the additional non-zero entry is for the state to remain unchanged. All other entries in a row are structural zeroes because those network states cannot be reached from the current state in a single change.

>>>>

<<child-demo, child='paper-child.Rnw'>>=
@
